A general and efficient representation of Ancestral Recombination Graphs

### 2022-07-19 paper.tex
Remove stochastic processes section
Fixes #220

\section*{Ancestral graphs and stochastic processes}
The coalescent~\citep{kingman1982coalescent,kingman1982genealogy,
hudson1983testing, tajima1983evolutionary}
provides the theoretical underpinning for much of contemporary population genetics.
\citet{hudson1983properties,hudson1990gene} extended the coalescent to include
recombination
and the model has proved to be an especially powerful basis for
simulations~\citep{hudson2002generating,baumdicker2021efficient}.
\cite{kelleher2016efficient} provide a description of his algorithm along with
an efficient implementation, the details of which turn out to be highly relevant to
later sections of the current paper (see the ``\nameref{Ancestry_resolution}'' section).

Although Hudson's algorithm was accompanied by a graph
schematic~\citep[][Fig. 5]{hudson1983properties}, the term
``Ancestral Recombination Graph'' or ARG, was introduced later by
Griffiths~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral},
who provided a closely related and complementary treatment.
Where Hudson's algorithm focusses on efficient \emph{simulation}, the original
ARG literature is focused on mathematical results about the underlying stochastic
process. Griffiths and colleagues used the term
``ARG'' to refer to both a particular branching-coalescing stochastic process and
the graph structure produced by this process.
% but for clarity and following recent usage,
% we will use ARG to refer to a graph structure, and refer to the
% process as the Griffiths process or the coalescent with recombination
% \citep{minichiello2006mapping}.

% NOTE: it might be helpful to have a drawing of the ARG process
% in-progress here to help us talk about it. What is the full state
% that is remembered as we go through the process? It does help to think
% about the process as choosing edges to merge, not nodes.
% We could also refer to Fig.~\ref{fig-arg-data-structure}B if we want a simple picture
% to go with this
In the Griffiths formulation, each edge in the graph corresponds to an extant
lineage and nodes are events in the process (the initial $n$ leaf nodes are
``sampling'' events). Common ancestor events occur at rate $\binom{k}{2}$ when there
are $k$ lineages present. We choose two lineages (edges) uniformly, and merge them
into a common ancestor lineage. Recombination events happen at a rate of
$k \rho$. Here, we choose a lineage (edge) uniformly, and a
breakpoint $0 < x < m$ uniformly on its genome. We terminate the edge at a
node, record the breakpoint and start two new edges from this node. The process
then continues until there is only one lineage left (the Grand Most Recent
Common Ancestor, GMRCA), which is guaranteed to
happen in finite time because of the linear vs quadratic rates of branching
and coalescing. Later, we will show how the graph structure and breakpoints
associated with
recombination nodes provides sufficient information to later recover the marginal
trees and calculate sampling probabilities. However, this process traces out
a larger graph (the ``big'' ARG) than the Hudson algorithm (the ``little'' ARG; see the ``\nameref{Ancestry_resolution}'' section),
and is consequently considerably less efficient \citep{wiuf1999ancestry}.

Similar graph encodings have also been introduced to model other mechanisms
which change the local ancestral tree along the genome, such as gene
conversion~\citep{wiuf2000coalescent} and
horizontal gene transfer~\citep{baumdicker2014infinitely}.
[Mention the
SMC~\citep{mcvean2005approximating,marjoram2006fast}
and ClonalFrame~\citep{didelot2007inference} approximations somewhere.]

\section*{The Event ARG data structure}\label{eARG}
% When did we move from the ARG is a process to the ARG is a data structure?
% What was the historical progression from one to the other?

% NB: should this para go into the previous section? It would be
% clean to start the section with the subsequent para
% starting "The Griffiths ARG encoding is an elegant and mathematically economical description of the CwR"
Early work on ARG inference focused on the problem of
inferring parameters of the
stochastic process, where the ancestry is regarded as a
latent parameter to be averaged out
\citep[e.g.][]{griffiths1996ancestral,kuhner2000maximum, nielsen2000estimation,
fearnhead2001estimating}.
These methods met with limited success
because the state space of ARGs is overwhelmingly large and
lacks a simple geometry or neighbourhood structure for inference or
sampling methods to  exploit.
The focus subsequently shifted to the more
tractable---but still NP-hard~\citep{wang2001perfect}---problem of computing
the minimum number of recombinations required
to explain an input dataset [CITATIONS], and inferring the corresponding
ARG realisations~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
This highlighted the distinction between the ARG as a stochastic process
and the ARG as a \emph{data structure}~\citep{minichiello2006mapping},
since parsimonious ARGs have no immediate connection to
the stochastic process.

% What do we mean by an EARG, informally?
The Griffiths ARG encoding is an elegant and mathematically
economical description of the coalescent with recombination, whose
fundamental units are the  \emph{events} that correspond directly
to those that occur in the associated stochastic process. We will refer
to this ARG encoding as an ``Event ARG'', or eARG:
an example is given in Fig.~\ref{fig-arg-data-structure}B.
Each node in the graph corresponds to an event: therefore the graph
contains both recombination nodes (node \textsf{d}) and common ancestor
nodes (nodes \textsf{e}, \textsf{f}, and \textsf{g}) corresponding to their
respective event types. Edges represent lineages,
and those lineages are modified by the events that they pass through.

@@

>>>>

% \section*{Ancestral graphs and stochastic processes}
% The coalescent~\citep{kingman1982coalescent,kingman1982genealogy,
% hudson1983testing, tajima1983evolutionary}
% provides the theoretical underpinning for much of contemporary population genetics.
% \citet{hudson1983properties,hudson1990gene} extended the coalescent to include
% recombination
% and the model has proved to be an especially powerful basis for
% simulations~\citep{hudson2002generating,baumdicker2021efficient}.
% \cite{kelleher2016efficient} provide a description of his algorithm along with
% an efficient implementation, the details of which turn out to be highly relevant to
% later sections of the current paper (see the ``\nameref{Ancestry_resolution}'' section).

% Although Hudson's algorithm was accompanied by a graph
% schematic~\citep[][Fig. 5]{hudson1983properties}, % but for clarity and following recent usage,
% we will use ARG to refer to a graph structure, and refer to the
% process as the Griffiths process or the coalescent with recombination
% \citep{minichiello2006mapping}.

% NOTE: it might be helpful to have a drawing of the ARG process
% in-progress here to help us talk about it. What is the full state
% that is remembered as we go through the process? It does help to think
% about the process as choosing edges to merge, not nodes.
% We could also refer to Fig.~\ref{fig-arg-data-structure}B if we want a simple picture
% to go with this
% In the Griffiths formulation, each edge in the graph corresponds to an extant
% lineage and nodes are events in the process (the initial $n$ leaf nodes are
% ``sampling'' events). Common ancestor events occur at rate $\binom{k}{2}$ when there
% are $k$ lineages present. We choose two lineages (edges) uniformly, and merge them
% into a common ancestor lineage. Recombination events happen at a rate of
% $k \rho$. Here, we choose a lineage (edge) uniformly, and a
% breakpoint $0 < x < m$ uniformly on its genome. We terminate the edge at a
% node, record the breakpoint and start two new edges from this node. The process
% then continues until there is only one lineage left (the Grand Most Recent
% Common Ancestor, GMRCA), which is guaranteed to
% happen in finite time because of the linear vs quadratic rates of branching
% and coalescing. Later, we will show how the graph structure and breakpoints
% associated with
% recombination nodes provides sufficient information to later recover the marginal
% trees and calculate sampling probabilities. However, this process traces out
% a larger graph (the ``big'' ARG) than the Hudson algorithm (the ``little'' ARG; see the ``\nameref{Ancestry_resolution}'' section),
% and is consequently considerably less efficient \citep{wiuf1999ancestry}.

% Similar graph encodings have also been introduced to model other mechanisms
% which change the local ancestral tree along the genome, such as gene
% conversion~\citep{wiuf2000coalescent} and
% horizontal gene transfer~\citep{baumdicker2014infinitely}.
% [Mention the
% SMC~\citep{mcvean2005approximating,marjoram2006fast}
% and ClonalFrame~\citep{didelot2007inference} approximations somewhere.]

\section*{The Event ARG data structure}\label{eARG}
Initial work on \emph{simulating} ancestral graphs with recombination was undertaken by
\citet{hudson1983properties,hudson1990gene}, who extended the classical coalescent 
framework~\citep{kingman1982coalescent,kingman1982genealogy, hudson1983testing, tajima1983evolutionary}
to include recombination. Hudson's algorithm has led to powerful simulation
tools~\citep{hudson2002generating,kelleher2016efficient, baumdicker2021efficient}, and is
also relevant to efficient analysis, as we shall see later in the
``\nameref{Ancestry_resolution}'' section.

% When did we move from the ARG is a process to the ARG is a data structure?
% What was the historical progression from one to the other?

The precise term ``Ancestral Recombination Graph'' or ARG, was subsequently introduced by
Griffiths and colleagues~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
Rather than simulation, their work focused on mathematical results about the underlying branching-coalescing
stochastic process. However, some confusion has arisen from their use of the word ``ARG'' to refer to
two distinct concepts: the process itself, and the graph structure it generates.

%% NB: I (yan) have commented out the following section as the "latent parameter" quote is
%% used again in the "ARG reconstruction" section, and may be more detail than is required here
%% This also loses the description of ARG space being large etc.
%Early work on ARG inference focused on the problem of
%inferring parameters of the
%stochastic process, where the ancestry is regarded as a
%latent parameter to be averaged out
%\citep[e.g.][]{griffiths1996ancestral,kuhner2000maximum, nielsen2000estimation,
%fearnhead2001estimating}.
%These methods met with limited success
%because the state space of ARGs is overwhelmingly large and
%lacks a simple geometry or neighbourhood structure for inference or
%sampling methods to  exploit.
%The focus subsequently shifted to the more
%tractable---but still NP-hard~\citep{wang2001perfect}---problem of computing
%the minimum number of recombinations required
%to explain an input dataset [CITATIONS], and inferring the corresponding
%ARG realisations~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
%This highlighted the distinction between the ARG as a stochastic process
%and the ARG as a \emph{data structure}~\citep{minichiello2006mapping},
%since parsimonious ARGs have no immediate connection to
%the stochastic process.

The second concept, that of the ARG being a \emph{data structure} was adopted by later authors,
when focussing on the problem of inferring an ARG-like structure (or a minimum number of
recombinations [CITATIONS]) without assuming a specific stochastic
model, for example using parsimony~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
Following~\citet{minichiello2006mapping}, we adopt this convention, using ARG
to mean an ancestral graph structure, without reference to the process that generated it.

% What do we mean by an EARG, informally?
\citet{griffiths1996ancestral} introduced an encoding for the ARG structure, which
is elegant and mathematically economical. In this encoding, the
fundamental units correspond directly to \emph{events}
that occur in the underlying stochastic process they were modelling.
We will refer to this ARG encoding as an ``event ARG'', or eARG:
an example is given in Fig.~\ref{fig-arg-data-structure}B.
Since each node in the graph corresponds to an event, the graph
contains both recombination nodes (node \textsf{d}) and common ancestor
nodes (nodes \textsf{e}, \textsf{f}, and \textsf{g}) corresponding to their
respective event types. Edges represent lineages,
and those lineages are modified by the events that they pass through.

@@

\textcolor{red}{Discussion on ASG ported from earlier (removed) ``stochastic processes'' section}
